MSc Project - Modelling a Double-Barred Galaxy as a Double Binary

### 2017-09-13 workspace/DoubleBars/Final/MScProjectDoubleBars.tex
Done!

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage[skip=0pt]{caption}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{MSc Project - Modelling a Double-Barred Galaxy as a Double Binary}
\begin{document}
\maketitle
\begin{abstract}
A brief literature review of double-barred galaxies and N-body simulations followed by an attempt at modelling a double-barred galaxy using a 4-body simulation of a double binary.  The aim of the project was to
find orbital resonances which may be responsible for the stability of double-barred galaxies. A 1:21 resonance was found but this is not representative of the observed pattern speeds of nested bars.
\end{abstract}
\section{Introduction}
\subsection{Double-Barred Galaxies} 
Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs \citep{vaucouleurs}, however it was only in the 1990s that double-barred galaxies
were recognised as being a distinct category of galaxies. The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. He found that inner bars were about 12\% the size of the outer bars with
semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \citep{erwin2}.
\cite{erwin1} believe that inner bars probably rotate independently of their outer bars and suggest that they are relatively long-lived structures.
Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}. The median size ratio is $\approx0.12$ with an upper limit of $\approx0.25$ which is consistent with the theory that
inner bars cannot be too large without disrupting the orbit of the outer bar.

@@

To counteract the effect of the increased mass of the inner binary in Figs. \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
inner binary has been increased. This has a noticeable stabilising effect on the inner binary which has regained approximate circularity.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-5-004/Orbit.eps}
\caption{Configuration 4}
\label{fig:config4}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-5-004/Inner.eps}
\caption{Configuration 4 - Inner Bar}
\label{fig:config4i}
\end{figure}

In figs. \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
of the increased mass of the inner binary. In order for these systems to be relatively stable the initial velocity of the inner binary was reduced. At this point 
both the inner and outer binaries had started to separate and the inner binary became increasingly elliptical.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-55-004/Orbit.eps}
\caption{Configuration 5}
\label{fig:config5}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-55-004/Inner.eps}
\caption{Configuration 5 - Inner Bar}
\label{fig:config5i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-57-004/Orbit.eps}
\caption{Configuration 6}
\label{fig:config6}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-57-004/Inner.eps}
\caption{Configuration 6 - Inner Bar}
\label{fig:config6i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-3/Orbit.eps}
\caption{Configuration 7}
\label{fig:config7}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-3/Inner.eps}
\caption{Configuration 7 - Inner Bar}
\label{fig:config7i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-4/Orbit.eps}
\caption{Configuration 8}
\label{fig:config8}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-4/Inner.eps}
\caption{Configuration 8 - Inner Bar}
\label{fig:config8i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-35/Orbit.eps}
\caption{Configuration 9}
\label{fig:config9}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-35/Inner.eps}
\caption{Configuration 9 - Inner Bar}
\label{fig:config9i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/006-6-006-3/Orbit.eps}
\caption{Configuration 10}
\label{fig:config10}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/006-6-006-3/Inner.eps}
\caption{Configuration 10 - Inner Bar}
\label{fig:config10i}
\end{figure}

In order to bring the orbits back into approximate circularity it was necessary to again increase the velocity of the inner binary. This is illustrated
in Figs. \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/025-65-02/Orbit.eps}
\caption{Configuration 11}
\label{fig:config11}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/025-65-02/Inner.eps}
\caption{Configuration 11 - Inner Bar}
\label{fig:config11i}
\end{figure}

@@

Further reduction of the velocity of the inner binary allowed further increase of inner binary mass without losing stability, Figs. \ref{fig:config19} to \ref{fig:config21i}. The simulation is
now stable enough for the outer binary to complete one orbit and proceed with a second.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/06-8-06-25/Orbit.eps}
\caption{Configuration 19}
\label{fig:config19}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/06-8-06-25/Inner.eps}
\caption{Configuration 19 - Inner Bar}
\label{fig:config19i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/08-9-08-15/Orbit.eps}
\caption{Configuration 20}
\label{fig:config20}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/08-9-08-15/Inner.eps}
\caption{Configuration 20 - Inner Bar}
\label{fig:config20i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/09-95-09-1/Orbit.eps}
\caption{Configuration 21}
\label{fig:config21}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/09-95-09-1/Inner.eps}
\caption{Configuration 21 - Inner Bar}
\label{fig:config21i}
\end{figure}

@@

Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilising and the 
simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive step-size becomes too small and the integrator is effectively not making any
progress with the integration and so ends the simulation. The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises. 1:21 could therefore possibly be an orbital resonance. 
Attempts at modifying the initial conditions of configuration 32 even slightly resulted in chaotic systems. The concept of a resonance is very familiar to us all: pushing someone on a playground swing at its natural
frequency will rapidly increase the swing amplitude \citep{malhotra}. Resonances occur naturally in the Solar system and there are three main types of resonance: spin-orbit, secular and mean motion \citep{malhotra}.
The most obvious example of a spin-orbit resonance is the Earth/Moon system where the Moon is spin-locked and only one side of the Moon is visible from the Earth. Secular resonances occur where the precession of two
orbits is synchronised. However, the 1:21 resonance, is an example of a mean-motion resonance where the orbital period of two bodies are close to a ratio of small integer \citep{malhotra}. Orbital resonances
generally results in destabilising a system however occasionally they may be stable and remain in resonance.
In the standard model of double-barred galaxies \citep{macie7} it is assumed that resonant coupling between bars is present: CR of the inner bar coincides with the ILR of the outer bar. Although \cite{macie4} 
found that resonant coupling between the bars may not be the dominant factor reducing chaos in the system.

Unfortunately, a 1:21 resonance doesn't represent the observational data of double-barred galaxies. \cite{font} found that the inner bar rotates more rapidly than the outer bar by a factor between 3.3 and 3.6.
Whilst \cite{macie4} predicted values for the ratio of bar angular rates between 1.6 and 2.7. With these studies in mind it may be worthwhile exploring the double binary model further and searching for stable
resonances around the 1:2 or 1:3 parameter space - see next section.

\subsection{Future Work}
As this system has a multidimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
It may be possible for an automated process to automatically reject systems in which there are collisions, or if any of the bodies exit the system, or if the system fails 
to persist for a certain number of orbits. In this way, it may be possible to find further orbital resonances.

@@

\bibitem[Alvarez–Ramirez and Medina(2014)]{alvarez}Alvarez–Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014

@@

%\bibitem[Maciejewski(2002)]{macie1}Maciejewski, W., Constraints on nested bars – implications for gas inflow, 2002, arXiv:astro-ph/0202110v1
%\bibitem[Maciejewski(2003)]{macie2}Maciejewski, W., Chaos or Order in Double Barred Galaxies?, 2003, arXiv:astro-ph/0304432v1

@@

>>>>

\documentclass[a4paper,12pt]{article}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{epsfig}
\usepackage{float}
\usepackage{natbib}
\usepackage{listings}
\usepackage[skip=0pt]{caption}
\graphicspath{{./}}
\DeclareGraphicsExtensions{.eps}
\author{Howard Kinsman}
\title{MSc Project - Modelling a Double-Barred Galaxy as a Double Binary}
\begin{document}
\maketitle
\begin{abstract}
A brief literature review of double-barred galaxies and N-body simulations followed by an attempt at modelling a double-barred galaxy using a 4-body simulation of a double binary.  The aim of the project was to
find orbital resonances which may be responsible for the stability of double-barred galaxies and attempt at verifying the calculations carried out by \cite{garzon}. 
A 1:21 resonance was found but this is not representative of the observed pattern speeds of nested bars.
\end{abstract}
\section{Introduction}
\subsection{Double-Barred Galaxies} 
Double-barred galaxies, where a smaller inner bar is nested inside a larger outer bar, were first observed in 1975 by de Vaucouleurs \citep{vaucouleurs}, however it was only in the 1990s that double-barred galaxies
were recognised as being a distinct category of galaxies. The current estimates of double-barred galaxy frequencies are $\approx30\%$ of barred galaxies and $\approx20\%$ of all galaxies \citep{erwin3}.
\cite{erwin2} surveyed 38 barred galaxies and found a total of 10 double-barred galaxies with no preference for Hubble type. He found that inner bars were about 12\% the size of the outer bars with
semi-major axes from $\approx250$pc to 1 kpc. Secondary bars probably rotate independently of, but in the same direction as their associated primary bars \citep{erwin2}.
\cite{erwin1} believe that inner bars probably rotate independently of their outer bars and suggest that they are relatively long-lived structures.
Inner bar size has been found to correlate slightly with outer bar size \citep{erwin3}. The median size ratio is $\approx0.12$ with an upper limit of $\approx0.25$ which is consistent with the theory that
inner bars cannot be too large without disrupting the orbit of the outer bar.

@@

To counteract the effect of the increased mass of the inner binary in Figs. \ref{fig:config4} and \ref{fig:config4i} the initial separation of the 
inner binary has been increased by .003 to .004. This has a noticeable stabilising effect on the inner binary which has regained approximate circularity.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-5-004/Orbit.eps}
\caption{Configuration 4}
\label{fig:config4}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-5-004/Inner.eps}
\caption{Configuration 4 - Inner Bar}
\label{fig:config4i}
\end{figure}

In Figs. \ref{fig:config5} to \ref{fig:config10i} the initial velocity of the outer binary has also been gradually increased to also counteract the effect
of the increased mass of the inner binary. In order for these systems to be relatively stable the initial velocity of the inner binary was reduced. At this point 
both the inner and outer binaries had started to separate and the inner binary became increasingly elliptical.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-55-004/Orbit.eps}
\caption{Configuration 5}
\label{fig:config5}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-55-004/Inner.eps}
\caption{Configuration 5 - Inner Bar}
\label{fig:config5i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-57-004/Orbit.eps}
\caption{Configuration 6}
\label{fig:config6}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/004-57-004/Inner.eps}
\caption{Configuration 6 - Inner Bar}
\label{fig:config6i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-3/Orbit.eps}
\caption{Configuration 7}
\label{fig:config7}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-3/Inner.eps}
\caption{Configuration 7 - Inner Bar}
\label{fig:config7i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-4/Orbit.eps}
\caption{Configuration 8}
\label{fig:config8}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-4/Inner.eps}
\caption{Configuration 8 - Inner Bar}
\label{fig:config8i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-35/Orbit.eps}
\caption{Configuration 9}
\label{fig:config9}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/005-58-005-35/Inner.eps}
\caption{Configuration 9 - Inner Bar}
\label{fig:config9i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/006-6-006-3/Orbit.eps}
\caption{Configuration 10}
\label{fig:config10}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/006-6-006-3/Inner.eps}
\caption{Configuration 10 - Inner Bar}
\label{fig:config10i}
\end{figure}

In order to bring the orbits back into approximate circularity it was necessary to again increase the velocity of the inner binary. This is illustrated
in Figs. \ref{fig:config11} and \ref{fig:config11i}. The mass of the inner binary has also increased considerably from the previous configurations to .025.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/025-65-02/Orbit.eps}
\caption{Configuration 11}
\label{fig:config11}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2016results/025-65-02/Inner.eps}
\caption{Configuration 11 - Inner Bar}
\label{fig:config11i}
\end{figure}

@@

Further reduction of the velocity of the inner binary allowed further increase of inner binary mass without losing stability, Figs. \ref{fig:config19} to \ref{fig:config21i}. The simulation is
now stable enough for the outer binary to complete one orbit and proceed with a second. It was found that by increasing the mass and separation of the inner binary by .01, and increasing the 
velocity of the outer binary by .05, whilst steadily reducing the velocity of the inner binary resulted in the fairly stable configurations shown here.
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/06-8-06-25/Orbit.eps}
\caption{Configuration 19}
\label{fig:config19}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/06-8-06-25/Inner.eps}
\caption{Configuration 19 - Inner Bar}
\label{fig:config19i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/08-9-08-15/Orbit.eps}
\caption{Configuration 20}
\label{fig:config20}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/08-9-08-15/Inner.eps}
\caption{Configuration 20 - Inner Bar}
\label{fig:config20i}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/09-95-09-1/Orbit.eps}
\caption{Configuration 21}
\label{fig:config21}
\end{figure}
\begin{figure}[H]
\centering
\includegraphics[width=0.9\textwidth]{./2017results/09-95-09-1/Inner.eps}
\caption{Configuration 21 - Inner Bar}
\label{fig:config21i}
\end{figure}

@@

Configuration 32 was by far the longest running simulation showing almost three complete orbits of the outer binary and many orbits of the inner binary. It showed no signs of destabilising and the 
simulation only ended because of a `Step size underflow' in the integrator. This error occurs when the adaptive step-size becomes too small and the integrator is effectively not making any
progress with the integration and so ends the simulation. The data shows that as the periodic ratios tends towards 1:21 the system considerably stabilises. 1:21 could therefore possibly be an orbital resonance. 
Attempts at modifying the initial conditions of configuration 32 even slightly resulted in chaotic systems. The concept of a resonance is very familiar to us all: pushing someone on a playground swing at its natural
frequency will rapidly increase the swing amplitude \citep{malhotra}. Resonances occur naturally in the Solar system and there are three main types of resonance: spin-orbit, secular and mean motion \citep{malhotra}.
The most obvious example of a spin-orbit resonance is the Earth/Moon system where the Moon is spin-locked and only one side of the Moon is visible from the Earth. Secular resonances occur where the precession of two
orbits are synchronised. However, the 1:21 resonance, is an example of a mean-motion resonance where the orbital period of two bodies are close to a ratio of small integer \citep{malhotra}. Orbital resonances
generally results in destabilising a system however occasionally they may be stable and remain in resonance.
In the standard model of double-barred galaxies \citep{macie7} it is assumed that resonant coupling between bars is present: CR of the inner bar coincides with the ILR of the outer bar. Although \cite{macie4} 
found that resonant coupling between the bars may not be the dominant factor reducing chaos in the system.

Unfortunately, a 1:21 resonance doesn't represent the observational data of double-barred galaxies. \cite{font} found that the inner bar rotates more rapidly than the outer bar by a factor between 3.3 and 3.6.
Whilst \cite{macie4} predicted values for the ratio of bar angular rates between 1.6 and 2.7. With these studies in mind it may be worthwhile exploring the double binary model further and searching for stable
resonances around the 1:2 or 1:3 parameter space - see next section. A comparison with the calculations carried out by \cite{garzon} was not completed because the results here seem to be dominated by free
oscillations rather than forced oscillations generated by the inner bar \citep{macie1}.

\subsection{Future Work}
As time constraints prevented the Garzon comparison from being carried out, a future project would need to minimise the free oscillations in order to compare it with Garzon's prediction.
As this system has a multidimensional parameter space, finding stable configurations was very time consuming. I suggest further studies in this area try to automate this process.
It may be possible for an automated process to automatically reject systems in which there are collisions, or if any of the bodies exit the system, or if the system fails 
to persist for a certain number of orbits. In this way, it may be possible to find further orbital resonances.

@@

\bibitem[Alvarez-Ramirez and Medina(2014)]{alvarez}Alvarez-Ramirez, M., Medina, M., A Review of the Planar Caledonian Four-Body Problem, 2014

@@

@@

\bibitem[Maciejewski(2017)]{macie1}Maciejewski, W., 11th September 2017 [email]
